---
title: "Lab 2: Presidents"
author: "Celeste Connell"
date: "9/6/2018"
output: pdf_document
subtitle: Sampling, plotting, descriptive statistics
---

# General information
This lab is due September 12th by 11:59 pm. You must upload your .rmd file and your knitted PDF to the dropbox on D2L. This lab is worth 10 points. You are welcome and encouraged to talk with classmates and ask for help. However, each student should turn in their own lab assignment and all answers, including all code, needs to be solely your own.

**\textcolor{blue} {This lab will make a lot more sense if you have already read Whitlock and Schluter, chapter 4, and are familiar with the concept of a sampling distribution. So, I strongly suggest reading through the chapter before doing this lab.}**

# Objective
The goal of this lab is to explore how we describe data (also known as descriptive statistics), as well as how we quantify uncertainty in data. You will also do some simulations regarding sampling to look at how sample sizes affect measures of uncertainty in data.

We will also continue to build your R skillset, including how to find new commands and manipulate datasets in R.

# Presidents dataset: working with data frames
Remember the presidential dataset we gathered last week in class? Read in the data from the .csv file on D2L. The data contains a list of presidents generated by our class, with each student naming a presidents. I have added their heights to this dataset. *IMPORTANT: When you read in your dataset, name the object `President.Sample`*. Your data should have only two columns, the first should have the presidents name (Last, First) and the second should have their height in cm.

\textcolor{red}{\textbf{Question 1 (0.25 points)}: Read in your data and save it as an object named `President.Sample`. Print out the first few rows using the $head()$ function. Make sure your data looks correct and has header labels.}
```{r}
rm(list=ls()) # clear previous variables
# start by setting your working directory (you can use code here, or use the 'files' window to the right)
setwd("~/College/Senior Year/Fall 2018/Biometry/Labs/Lab 2")
# read in your data and name it President.Sample
President.Sample <- read.csv("SamplePres.csv")
# print off the first few rows
head(President.Sample)
```


Although excel can be a good way of organizing your data, learning how to manipulate your data within R is an important skill, and one we'll work on throughout the semester. Let's go through a few common tasks you may encounter while organizing your data.

## Adding and removing columns
In this example code, I will add a column, called `Political_Party`, and then remove it.

First, we will add a column called "Political_Party" that is filled with blanks (NAs):
```{r}
President.Sample$Political_Party <- NA
head(President.Sample)
```
In this code, I use `<-NA` to create a column filled with NAs. NA is what R uses to code a blank cell. Now that the column is created, I could then enter in the political parties in one by one. Let's do the 22nd president in the dataset (Lincoln)

```{r}
President.Sample$Political_Party[22] <- "Republican"
```

Here I took the 22nd entry of the `Political_Party` column, and gave it the value `Republican`. Note that I put the word in quotes. If it were a number instead of text, I wouldn't need to use quotes. 

If we decide we don't want this column anymore, we can delete it:
```{r}
President.Sample$Political_Party <- NULL
```

To make sure the column was actually deleted, you can print out the header again, or you can just view the column names to make sure it isn't there.
```{r}
colnames(President.Sample)
```

Great! `Political_Party` is no longer a column in the dataset.

Another quick way to look at your data is by going to the "environment" tab in Rstudio and finding your dataframe under the Data section. Double click on your dataframe name. It will open in a new window, letting you look at it as if it were an excel spreadsheet. You can click on header names to sort by columns, too. This is a nice way to scan your dataset for errors or make sure you haven't accidentally deleted anything while coding.

Now, you should add, and then delete, a column to our dataframe, called "gender". You can use a similar approach as I have outlined above.

\textcolor{red}{\textbf{Question 2 (0.25 points)}: Add, and then delete, a column to your dataframe which gives the gender of each president. Note that when you make character vectors (as opposed to numeric vectors), each element must be enclosed in a set of quotations, i.e. "female" instead of female. (Hint: due to our unfortunate lack of presidential diversity, you can use a similar strategy as above, except instead of creating a column of NAs, you would create a column of "Male" entries.)}

```{r}
# add a new column, called "gender", with the gender of each president
President.Sample$gender <- "male"
head(President.Sample)
# now delete it
President.Sample$gender <- NULL
colnames(President.Sample)
```

## Performing calculations on your data 
R can easily perform calculations on entire vectors or columns, of data. For instance:
```{r}
myvector<-seq(0,10,1)
myvector2<-myvector-1
myvector
myvector2
```

Print out myvector and myvector2 in your console. How was the -1 calculation applied? 

\textcolor{red}{\textbf{Question 3 (0.5 points)}: create a new column, giving the heights of each president in inches rather than cm. Name this column "heightin"}
```{r}
President.Sample$heightin <- President.Sample$Height*0.393701
President.Sample
```


## Data clean up
Data clean up can be the most challenging part of a project, especially when working with large datasets. Common problems include missing and duplicate data. Do we have duplicates in our president dataset? Run each line of code below and view the output to see what each line of code does.
```{r}
# This code finds anywhere where the entire row is duplicated
duplicated(President.Sample)
```

Each entry in this TRUE/FALSE vector represents a row of the data. A TRUE means that row is a complete duplicate of another row.

```{r}
# This code finds duplicates where an entry in one column (i.e. just the name) is duplicated
duplicated(President.Sample$Name)
```

This would allow us to find where we have the same president in our dataset more than once, but the heights don't match exactly.

```{r}
# This code prints out a dataframe of just the unique data
unique(President.Sample)
```

Notice that the original row number is preserved, but the dataframe is much smaller than it was originally. You can use `length()`, `str()`, or `nrow()` functions to convince yourself of this.

We might want to save the unique records as a new dataframe to work with:
```{r}
Pres.Samp.2<-unique(President.Sample)
```

# Descriptive statistics

**Important!** From this point forward, use your `Pres.Samp.2` dataset, which has the duplicates removed. 

R has several built-in functions to calculate mean (*mean*), standard deviation (*sd*), minimum (*min*), and maximum (*max*) of a vector. We'll be using these a lot! 

\textcolor{red}{\textbf{Question 4 (1 point)}: From your dataset, calculate the mean, median, standard deviation, minimum, and maximum of presidential heights (in inches). Save each of these numbers as a different object so you can look them up later. Show your code.}

```{r}
meanheight.Pres.Samp.2 <- mean(Pres.Samp.2$heightin)
medianheight.Pres.Samp.2 <- median(Pres.Samp.2$heightin)
sdheight.Pres.Samp.2 <- sd(Pres.Samp.2$heightin)
minheight.Pres.Samp.2 <- min(Pres.Samp.2$heightin)
maxheight.Pres.Samp.2 <- max(Pres.Samp.2$heightin)
```


What if we want to know something specific about our dataset, for example: who is the tallest president? We can use the *which* argument in R. 
```{r}
tallest<-which(Pres.Samp.2$Height==max(Pres.Samp.2$Height))
tallest
```

Note that if you print the object "tallest", you get a number, or multiple numbers if two presidents are "tied" for first place. The `which()` function is a handy way to get the index (i.e. row number) that contains the maximum height value. The double equal sign in `Pres.Samp.2$heightin==max(Pres.Samp.2$heightin)` tests if two things are equal. In this case we are testing if the height in each row equals the maximum height. This produces true or false values as you can see below.

```{r}
Pres.Samp.2$Height==max(Pres.Samp.2$heightin)
```

Then we use the `which()` function, which, when wrapped around the above code extracts the index number of just the TRUE values. If more than one president is tied for first, it returns multiple row indices. Both the double equals and the `which()` function are used a lot in data analysis and programing. Similar to the double equals, you can also use <=, <, >, or >= to find all of the elements less than (or, correspondingly, more than) a given value.

So, we've gotten the row number(s) which contains the tallest president(s), but not the name(s) of the tallest president(s). Let's print out the whole row so we can see who it is. By now, these square brackets should look familiar. We are telling R we want just the row with the tallest president, and we want all the columns.

```{r}
Pres.Samp.2[tallest,]
```

\textcolor{red}{\textbf{Question 5 (0.5 points)}: Who is the shortest president in our dataset? Modify the code demonstrated above to print out the row containing the shortest president, and write his name and height below.}

```{r}
shortest<-which(Pres.Samp.2$Height==min(Pres.Samp.2$Height))
shortest
Pres.Samp.2[shortest,]
```

*John Adams is the shortest president, at 170 cm or 66.92917 inches* 

The `which()` function is not the most intuitive, so don't worry too much if you do not completely understand how to use it. It is a great tool for extracting data but it's not crucial for the rest of the lab assignment. I just want you to get used to seeing it.

# Comparing sample means to true means
Last week, we "randomly" sampled presidents and measured (via Wikipedia) their heights. Suppose we wanted to use this dataset to estimate the average height of all American presidents. 

\textcolor{red}{\textbf{Question 6 (0.5 point)}: Is our class sample a random sample? Why or why not?}

*Our class sample is not a random sample. We're likely to be biased towards the most recent or iconic presidents.*


\textcolor{red}{\textbf{Question 7 (0.5 point)}: Are our data points independent? Why or why not?}

*No, they're not independent. Since each student wrote the name of their president on the board, it's likely that this visual factor influenced the decisions (consciously or subconsciously) of students who hadn't wrote theirs on the board yet.*



\textcolor{red}{\textbf{Question 8 (0.5 point)}: How would we find the true parameter for the variable we are estimating?}

*We'd have to find a list of every American president*



Now let's compare our sample mean (our estimate) to the true population mean (the parameter). Luckily the true population of American presidents is finite, and quite easy to find (for most ecological data we have little hope of ever measuring all of the units in our population).

**Read in the dataset, "presidential_heights.csv", posted to D2L**. Make sure you name the dataframe something distinct from your class dataset.
```{r}
# read in the .csv file containing all the presidents
President.Heights <- read.csv("presidential_heights.csv")
# explore your data: how many rows, columns? What type of data does it include?
class(President.Heights)
names(President.Heights)
str(President.Heights)
```
*Answer: A data frame with 45 rows and 3 columns. The first column "Name" is a factor variable; second and third columns "Height" and "InaugurationYr" are integer variables.*

Before we compare our sample of presidents to the real population of presidents, let's practice making some histograms.

##Histograms
Histograms are a great way to visualize raw data, and we will use them a lot over the course of the semester. The command `hist()` plots a histogram. Remember that a histogram plots the frequency of one variable, so we want to make sure to give the command a vector, rather than a whole dataframe. Explore the code below which uses R's example CO2 dataset. There are many optional arguments we can add to `hist()` to make the histogram more appealing to the eye.  In R markdown the figures appear below the code chunk, rather than in the plots window to the right. If you want to close the plot, click the tiny gray X in the upper right corner. Explore and practice editing the code below to understand some of the basic modifications we can make. (You do not need to write the answers down anywhere; this is for your own understanding.)

```{r, fig.width=3, fig.height=3}
# the most basic histogram:
hist(CO2$uptake)
```

Now, let's try a few modifications:
```{r, fig.width=3, fig.height=3}
hist(CO2$uptake, breaks=20) 
```

In the above code, try changing 20 to something else. What does it do?

```{r, fig.width=3, fig.height=3}
hist(CO2$uptake, main="Poo", xlim=c(-50,100))
```

In the above code, change the -10 to -50 and the 60 to 100. What happens?

Change the "Carbon Uptake" to some other text. What happens?

```{r, fig.width=3, fig.height=3}
hist(CO2$uptake, xlab="uptake !", ylab="count!",col="indianred2") 
```

Practice changing the "uptake", "count", and color arguments. 


The code below makes a panel of two histograms, with the top showing uptake, and the bottom showing concentrations. If you want the two histograms to be plotted at once, you must enclose all the plot commands inside curly brackets `{}`.

```{r, fig.width=5, fig.height=5}
{par(mfrow=c(2,1))
hist(CO2$uptake, main="Carbon Uptake")
hist(CO2$conc,main="Concentration")
par(mfrow=c(1,1))} #set the plotting back to a single histogram per plot
```

The command `par`, used above, allows you to set a number of graphical parameters, including how plots are arranged. We will explore its capabilities further as the semester progresses. The `mfrow=c(2,1)` argument tells R you want 2 rows of plots and 1 column, whereas `mfrow = c(1,2)` would make 2 columns of plots but only one row. Therefore, the latter code would make a side by side rather than top/bottom plot. You can do more than 2 plots per panel, i.e. `mfrow=c(3,2)`, however, you will have to modify your text and plot sizes to make sure they fit, and this can be a pain. 

\textcolor{red}{\textbf{Question 9 (1 point):} Now visually compare the height (in inches) of sampled presidents vs. the true population. Create a two paneled figure with a histogram of the sampled presidents on the top row and full president population on the bottom row. To better visually compare the distributions, make sure they have the same xlim scale. You may want to modify things like: number of breaks, label axes, titles, or color as shown above.}

```{r,fig.width=6, fig.height=6}
{par(mfrow=c(2,1))
hist(Pres.Samp.2$heightin, main="President Heights (Sample)", xlab="Heights in inches", col="red", breaks=6)
hist(President.Heights$Height,main="President Heights (Actual)", xlab="Heights in inches", col="blue", breaks=6)
par(mfrow=c(1,1))} 
```



\textcolor{red}{\textbf{Question 10 (1 point):} Calculate the mean, median, and standard deviation of height in the full presidents dataset. Compare these to the estimates you found in Question 4.}

```{r}
meanheight.Pres <- mean(President.Heights$Height)
medianheight.Pres <- median(President.Heights$Height)
sdheight.Pres <- sd(President.Heights$Height)
meanheight.Pres
meanheight.Pres.Samp.2
medianheight.Pres
medianheight.Pres.Samp.2
sdheight.Pres
sdheight.Pres.Samp.2

```


# Sampling Distributions
What if we asked 1000 different biometry classes to do the same thing we did and get a sample of presidential heights? How would all of our samples compare to one another, and how would they compare to the true population distribution of presidents? 

We will simulate this data below. Run the code chunk below, and, if you like, explore the output of each line. *Don't worry if the loop coding is confusing; we will cover loops later on in the semester!*

The code below simulates 1000 different sample sets, all of which are the same size as our class' sample set (n = 12, without the duplicates). We'll draw a random sample of 12 from the full list of 44 presidents, then repeat 1000 times. **Note that this truly random sampling is different than how we sampled last week--more on that later.**

Make a mental note that we aren't drawing 1000 random presidents, but rather 1000 different *datasets*, each one containing 12 randomly chosen presidents. What we are doing here is simulating the *sampling distribution of the means*. Don't worry too much about how the code below works, but make sure you run it.

```{r SIMULATED DATA}
FullPres.Heights <- c(74,67,75,64,72,67,73,66,68,72,68,68,69,70,72,74,70,68,68,72,74,66,
                      71,67,70,72,71,72,70,71,74,69,70,72,75,72,72,69,74,74,74,71,73,74)
# First, let's simulate one randomly sampled dataset of 12 presidential heights
random.indices<-sample(1:length(FullPres.Heights), 12, replace=FALSE)
simulatedsample1<-FullPres.Heights[random.indices]
simulatedmean1<-mean(simulatedsample1); simulatedsd1<-sd(simulatedsample1)
# Now, let's repeat this process to generate 1000 random sample sets
# then we can compute the mean of each dataset for our sampling distribution
# make an empty vector which we will fill with our simulated means:
simulatedmeans<-rep(NA,1000) 
# repeat the process we did above 1000 times
# each time we put the mean of the random sample we drew into our empty vectors
for (i in 1:1000){
  random.indices<-sample(1:length(FullPres.Heights), 14, replace=FALSE)
  sim.sample<-FullPres.Heights[random.indices]
  sim.mean<-mean(sim.sample)
  simulatedmeans[i]<-sim.mean
  }
```

Phew! That was a lot of code. But, now we have a vector `simulatedmeans` which contains the \textbf{Sampling Distribution of the Mean}. Run the code below to take a look at the shape of this distribution:

```{r PLOT SIMULATED DATA}
{hist(simulatedmeans, main="Sampling Distribution", 
      xlab="Mean Height (inches) of sample datasets", cex=.8, breaks=15,
      col="gray", xlim = c(65,75))
# draw a green line where the mean of all the 1000 samples would fall
abline(v = mean(simulatedmeans), col="green", lwd=3)  
# draw a blue line where our class data would fall
abline(v=mean(Pres.Samp.2$heightin), col="blue", lwd=2)
# draw a red line where the population mean (mu) falls
abline(v=70.78, col="red", lwd=2)}
```

How well would any random sample of 12 presidents (which was the size of our class dataset after subtracting duplicates) work to estimate the height parameter? The mean of all of our SIMULATED data (green) is really close to the mean of the full presidents dataset, or the true population parameter (in red). That's good news, and is what we expect if we are drawing a truly random sample. If the peak of our histogram is close to the true population parameter we likely have an unbiased way of collecting data. If we did 10,000 random datasets, our histogram would look even nicer.

Most of the sample means are pretty close to the true parameter, but notice that are some sample sets that are especially low or high (near the tails of the histogram). Some random samples of 12 presidents produce very poor estimates of the population parameter, but those samples are rare.

\textcolor{red}{\textbf{Question 11 (1 point):} Where does our class estimate, in blue, fall on this distribution? Based on how we sampled the presidents in our class, why is our estimate atypical? Give a reason to explain why our sampling led to an estimate that is far from the true parameter.}

*Our class estimate falls very far right from the population mean. It looks like 11 out of our 12 sampled presidents have heights above the population mean of 70.77, leading to a high sample mean on our distribution.*


# Calculating descriptive statistics and plotting: tree data
The dataset `trees` gives the girth, height, and volume of a few dozen randomly sampled black cherry trees. Suppose we want to use this data to estimate the true mean girth of black cherry trees. 

```{r}
# preview the data
head(trees)
```

## Sampling error
If we want to know how accurate the estimate of mean girth is, we should calculate the sample's *standard error*. 

The approximated standard error of the mean is called $SE_{\bar{Y}}$, commonly shortened as SE. It is calculated as the sample standard deviation *s* divided by the square root of the sample size.

\[SE = s / \sqrt(n)\]

(Note that this equation will look funny to you when viewing in R markdown, but will be formatted nicely in the PDF!)

Remember, to calcualate standard error, you will use the *sample* standard deviation (s) and not the full *population* standard deviation, ($\sigma$), which is a parameter we usually don't know.

\textcolor{red}{\textbf{Question 12 (1 point):} Using functions in R, calculate the mean, standard deviation, and standard error of black cherry tree girth from the sample data in `trees`. Save them as variables and print them. Hint: use the length() or nrow() function to find the sample size of this dataset.}

```{r}
nrow(trees)
mean.trees <- mean(trees$Girth)
mean.trees
sd.trees <- sd(trees$Girth)
sd.trees
se.trees <- (sd.trees / sqrt(31))
se.trees
```


\textcolor{red}{\textbf{Question 13 (1 point):} How would increasing the sample size change our estimate of standard error? How would increasing the sample size change our estimate of the mean girth (x-bar)?}

*Increasing the sample size would decrease our estimate of the standard error. Also, increasing the sample size wouldn't affect our sample estimate of mean girth.* 


## Graphics practice

\textcolor{red}{\textbf{Question 14 (0.5 point):} Create a scatterplot showing the relationship between girth and height of cherry trees. Change at least 2 of the defaults to improve the aesthetics of your plot. Make sure your plot has informative axis labels (no dollar signs!).}

```{r}
plot(trees$Girth, trees$Height, xlab="Tree Girth", ylab="Tree Height", main="Cherry Tree Girth vs. Height", col="mediumvioletred", pch=11)
?plot
```


# Practice with stripcharts: Iris example
The `iris` dataset in R contains flower characteristics for multiple species of iris. 

```{r}
head(iris)
```

\textcolor{red}{\textbf{Question 15 (0.5 point):} Create a stripchart that shows the differences in petal width between species (hint: see lecture 4). For a *bonus* point of 1, add means and standard deviation bars.}

```{r}
#stripchart
stripchart(Petal.Width ~ Species, data=iris, vertical=TRUE, method="jitter", 
           xlab="Species", ylab="Petal Width",
           pch=16, col="lightpink4", main="Differences in petal width between iris species")

#calculate means and sd
mean.iris <- tapply(iris$Petal.Width, iris$Species, mean)
sds.iris <- tapply(iris$Petal.Width, iris$Species, sd)

#add means and sd to graph
segments(x0=c(1:3), x1=c(1:3),y0=mean.iris-sds.iris,y1=mean.iris+sds.iris,lwd=2, col="forestgreen")
points(x=c(1:3), y=mean.iris,pch=19, col="yellow")
```

